---
title: "Final Project---Unemployment During Presidential Terms"
author: "Chang, Emily' Davis, Alexander; Henry, Colin"
date: 'Due: Wed Dec 4 | 11:59pm'
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

**APMA 3150 \| Fall 2024 \| University of Virginia**

***"On my honor, I pledge that I have neither given nor received unauthorized aid on this assignment." - Emily Chang.***

```{r}
library(dplyr)
```

# Preprocessing
```{r}
natl_unemploy <- read.csv('data/national_unemployment_rate.csv')
presidents <- read.csv('data/presidents.csv')

natl_unemploy$DATE <- as.Date(natl_unemploy$DATE)

presidents$Party <- trimws(presidents$Party)
presidents$Party <- as.factor(presidents$Party)
presidents$Took.office <- as.Date(presidents$Took.office, format = "%d/%m/%Y")
presidents$Left.office <- c(presidents$Took.office[-1] - 1, as.Date('9999-12-31'))

natl_unemploy_pres <- natl_unemploy %>%
  left_join(presidents, by = character()) %>%
  filter(DATE >= Took.office & DATE <= Left.office)

natl_unemploy_pres <- natl_unemploy_pres %>%
  select(DATE, UNRATE, President, Party)
head(natl_unemploy_pres)
```

# Test for Normality
This data is not normally distributed
```{r}
hist(natl_unemploy_pres$UNRATE, main="Histogram of Unemployment Rates", xlab="Unemployment Rate", col="lightblue", border="black")
qqnorm(natl_unemploy_pres$UNRATE, main="Q-Q Plot of Unemployment Rate", 
       xlab="Theoretical Quantiles", ylab="Sample Quantiles")
qqline(natl_unemploy_pres$UNRATE, col = "red")
```

# Testing for Statistical Significance Between Political Parties
## Two-tailed
In this case:

* $H_0$:  The null hypothesis is that the median of the population of differences between the paired data is zero.
* #H_A$: The alternative hypothesis is that it is not.

Under this hypothesis test, we can reject the null hypothesis at the 95% confidence level. 
```{r}
wilcox.test(UNRATE ~ Party, data = natl_unemploy_pres, alternative='two.sided')
```

## One-tailed
### alternative=greater
In this case:

* $H_0$:  The null hypothesis is that the median of the population of differences between the paired data is zero.
* $H_A$: The alternative hypothesis is that the median of the Democratic population is greater than that of the Republicans.

Under this hypotbesis test, we fail to reject the null hypothesis at the 95% confidence level.
```{r}
wilcox.test(UNRATE ~ Party, data = natl_unemploy_pres, alternative='greater')
```
### alternative=less
In this case:

* $H_0$:  The null hypothesis is that the median of the population of differences between the paired data is zero.
* $H_A$: The alternative hypothesis is that the median of the Democratic population is less than that of the Republicans.

Ybder this hypothesis test, we can reject the null hypothesis at the 95% confidence level.
```{r}
wilcox.test(UNRATE ~ Party, data = natl_unemploy_pres, alternative='less')
```