---
title: "Final Project---Unemployment During Presidential Terms"
author: "Chang, Emily' Davis, Alexander; Henry, Colin"
date: 'Due: Wed Dec 4 | 11:59pm'
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

**APMA 3150 \| Fall 2024 \| University of Virginia**

***"On my honor, I pledge that I have neither given nor received unauthorized aid on this assignment." - Emily Chang.***

```{r}
library(dplyr)
library(ggplot2)
```

# Part 1
## Preprocessing
```{r}
natl_unemploy <- read.csv('data/national_unemployment_rate.csv')
presidents <- read.csv('data/presidents.csv')

natl_unemploy$DATE <- as.Date(natl_unemploy$DATE)

presidents$Party <- trimws(presidents$Party)
presidents$Party <- as.factor(presidents$Party)
presidents$Took.office <- as.Date(presidents$Took.office, format = "%d/%m/%Y")
presidents$Left.office <- c(presidents$Took.office[-1] - 1, as.Date('2017-01-20'))

natl_unemploy_pres <- natl_unemploy %>%
  left_join(presidents, by = character()) %>%
  filter(DATE >= Took.office & DATE <= Left.office)

natl_unemploy_pres <- natl_unemploy_pres %>%
  select(DATE, UNRATE, President, Party)
head(natl_unemploy_pres)
natl_unemploy_pres$Party <- factor(natl_unemploy_pres$Party, levels = c("Democratic", "Republican"))
```

## Test for Normality
This data is somewhat normally distributed
```{r}
hist(natl_unemploy_pres$UNRATE, main="Histogram of Unemployment Rates", xlab="Unemployment Rate", col="lightblue", border="black")
qqnorm(natl_unemploy_pres$UNRATE, main="Q-Q Plot of Unemployment Rate", 
       xlab="Theoretical Quantiles", ylab="Sample Quantiles")
qqline(natl_unemploy_pres$UNRATE, col = "red")
```
```{r}
shapiro.test(natl_unemploy_pres$UNRATE)
```
This data is not normally-distributed as the p-value from the Shapiro-Wilks test is very low.

## Testing for Statistical Significance Between Political Parties
### Wilcoxon

```{r}
median_values <- natl_unemploy_pres %>%
  group_by(Party) %>%
  summarise(Median = median(UNRATE, na.rm = TRUE))

median_values

ggplot(natl_unemploy_pres, aes(x = UNRATE, color = Party)) +
  geom_density(size = 1) +
  geom_vline(data = median_values, aes(xintercept = Median, color = Party), 
             linetype = "dashed", size = 1.2) +
  labs(title = "Density Plot of Unemployment Rates by Party",
       x = "Unemployment Rate (%)", y = "Density") +
  scale_color_manual(values = c("blue", "red")) +
  theme(
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  )
```

#### Two-tailed
In this case:

* $H_0$:  The null hypothesis is that the mean of the population of differences between the paired data is zero.
* $H_A$: The alternative hypothesis is that it is not.
```{r}
dem_rows <- subset(natl_unemploy_pres, Party == "Democratic")
dem_unemploy_avg <- mean(dem_rows$UNRATE)
paste("Average unemployment rate during a Democratic presidential term:", dem_unemploy_avg)

rep_rows <- subset(natl_unemploy_pres, Party == "Republican")
rep_unemploy_avg <- mean(rep_rows$UNRATE)
paste("Average unemployment rate during a Republican presidential term:", rep_unemploy_avg)
```
Under this hypothesis test, we can reject the null hypothesis at the 95% confidence level.
$H_0$: the two populations have the same distribution with the same median
$H_A$: the two populations have distributions and medians that differ significantly.
```{r}
wilcox.test(UNRATE ~ Party, data = natl_unemploy_pres, alternative='two.sided')
```

#### One-tailed
##### alternative=greater
In this case:

* $H_0$: the two populations have the same distribution with the same median
* $H_A$: the two populations do not have the same distribution and the median of the Democratic population is greater than that of the Republicans.

Under this hypotbesis test, we fail to reject the null hypothesis at the 95% confidence level.
```{r}
wilcox.test(UNRATE ~ Party, data = natl_unemploy_pres, alternative='greater')
```
##### alternative=less
In this case:

* $H_0$: the two populations have the same distribution with the same median
* $H_A$: the two populations do not have the same distribution and the median of the Democratic population is less than that of the Republicans.

Under this hypothesis test, we can reject the null hypothesis at the 95% confidence level.
```{r}
wilcox.test(UNRATE ~ Party, data = natl_unemploy_pres, alternative='less')
```

### t-test

* $H_0: \mu_{Democratic} = \mu_{Republican}$
* $H_A: \mu_{Democratic} \neq \mu_{Republican}$

```{r}
levels(natl_unemploy_pres$Party)
print(t.test(UNRATE ~ Party, data = natl_unemploy_pres))
```
Because the p-value is less than 0.05, we can reject the null hypothesis at the 95% confidence level.

```{r}
print(t.test(UNRATE ~ Party, data = natl_unemploy_pres, alternative = "greater"))
```
Due to the high p-value, We fail to reject the null hypothesis when the alternative hypothesis is set "true difference in means between group Democratic and group Republican is greater than 0."

```{r}
print(t.test(UNRATE ~ Party, data = natl_unemploy_pres, alternative = "less"))
```
Due to the low p-value, we reject the null hypothesis when the alternative hypothesis is set "true difference in means between group Democratic and group Republican is less than 0."

# Part 2
## Preprocessing
```{r}
num_of_data_points_per_president <- natl_unemploy_pres %>%
  group_by(President) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))
num_of_data_points_per_president
```

The number of data points for each presidential term ranges from 29 to 189. Because there does not exist a presidential term with a large amount of data points, we use the t-test to test for statistical significance between one one presidential term and another.

## T-Test
* $H_0: \mu_1 = \mu_2$ (where $\mu_1$ is the average monthly unemployment rate for the first president, and $\mu_2$ is the average monthly unemployment rate for the second president)
* $H_A: \mu_1 \neq \mu_2$
```{r}
pres_data <- natl_unemploy_pres %>%
  group_by(President) %>%
  summarise(
    Start = min(DATE),
    UnemploymentRates = list(UNRATE)
  ) %>%
  arrange(Start)

for (i in 1:(nrow(pres_data) - 1)) {
  president1 <- pres_data$President[i]
  president2 <- pres_data$President[i + 1]
  
  term1_data <- unlist(pres_data$UnemploymentRates[i])
  term2_data <- unlist(pres_data$UnemploymentRates[i + 1])
  mean_term1 <- mean(term1_data)
  mean_term2 <- mean(term2_data)
  
  cat("\nComparison:", president1, "vs", president2, "\n")
  cat("Sample mean of", president1, ":", mean_term1, "\n")
  cat("Sample mean of", president2, ":", mean_term2, "\n")
  cat("Two-tailed t-test:\n")
  print(t.test(term1_data, term2_data, paired = FALSE))
  cat("One-tailed t-test (less):\n")
  print(t.test(term1_data, term2_data, paired = FALSE, alternative = "less"))
  cat("One-tailed t-test (greater):\n")
  print(t.test(term1_data, term2_data, paired = FALSE, alternative = "greater"))
  cat("------------------------------------------------------------------------")
}
```
* Harry S. Truman VS Dwight D. Eisenhower
  * The sample average monthly unemployment rate of Truman is lower than that of Eisenhower.
  * The two-tailed t-test resulted in a p-value of 0.001303.
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Truman} < \mu_{Eisenhower}$ results in a p-values of 0.0006515.
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Truman and Eisenhower presidential terms.
  * There is also evidence that Truman's monthly average unemployment rates were lower than that of Eisenhower's.
  
* Dwight D. Eisenhower VS John F. Kennedy
  * The sample average monthly unemployment rate of Eisenhower is lower than that of Kennedy.
  * The two-tailed t-test resulted in a p-value of 1.369e-09
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Eisenhower} < \mu_{Kennedy}$ results in a p-values of 6.845e-10.
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Eisenhower and Kennedy presidential terms.
  * There is also evidence that Kennedy's monthly average unemployment rates were greater than that of Eisenhower's.
  
* John F. Kennedy VS Lyndon B. Johnson
  * The sample average monthly unemployment rate of Kennedy is greater than that of Johnson
  * The two-tailed t-test resulted in a p-value of 2.2e-16
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Kennedy} > \mu_{Johnson}$ results in a p-values of 2.2e-16.
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Kennedy and Johnson presidential terms.
  * There is also evidence that Kennedy's monthly average unemployment rates were greater than that of Johnson's.

* Lyndon B. Johnson VS Richard Nixon
  * The sample average monthly unemployment rate of Johnson is lower than that of Nixon
  * The two-tailed t-test resulted in a p-value of 1.733e-09
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Johnson} < \mu_{Nixon}$ results in a p-values of 8.666e-10.
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Johnson and Nixon presidential terms.
  * There is also evidence that Nixon's monthly average unemployment rates were greater than that of Johnson's.
  
* Richard Nixon VS Gerald Ford
  * The sample average monthly unemployment rate of Nixon is lower than that of Ford
  * The two-tailed t-test resulted in a p-value of 2.2e-16
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Nixon} < \mu_{Ford}$ results in a p-values of 2.2e-16.
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Nixon and Ford presidential terms.
  * There is also evidence that Nixon's monthly average unemployment rates were less than that of Johnson's.
  
* Gerald Ford VS Jimmy Carter
  * The sample average monthly unemployment rate of Ford is greater than that of Carter
  * The two-tailed t-test resulted in a p-value of 2.687e-10
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Ford} > \mu_{Carter}$ results in a p-values of 1.343e-10.
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Ford and Carter presidential terms.
  * There is also evidence that Ford's monthly average unemployment rates were greater than that of Carter's.
  
* Jimmy Carter VS Ronald Reagan
  * The sample average monthly unemployment rate of Carter is less than that of Reagan
  * The two-tailed t-test resulted in a p-value of 2.223e-07
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Carter} < \mu_{Reagan}$ results in a p-values of 1.112e-07.
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Carter and Reagan presidential terms.
  * There is also evidence that Carter's monthly average unemployment rates were less than that of Reagan's.
  
* Ronald Reagan VS George H. W. Bush
  * The sample average monthly unemployment rate of Reagan is greater than that of Bush
  * The two-tailed t-test resulted in a p-value of 5.356e-08
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Reagan} > \mu_{Bush}$ results in a p-values of 2.678e-08.
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Reagan and Bush presidential terms.
  * There is also evidence that Reagan's monthly average unemployment rates were greater than that of Bush's.
  
* George H. W. Bush VS Bill Clinton
  * The sample average monthly unemployment rate of Bush is greater than that of Clinton
  * The two-tailed t-test resulted in a p-value of 3.293e-10
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Bush} > \mu_{Clinton}$ results in a p-values of 1.646e-10.
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Bush and Clinton presidential terms.
  * There is also evidence that Bush's monthly average unemployment rates were greater than that of Clinton's.
  
* Bill Clinton VS George W. BUsh
  * The sample average monthly unemployment rate of Bush is greater than that of Clinton
  * The two-tailed t-test resulted in a p-value of 0.2505
  * There does not exist enough evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Bush and Clinton presidential terms.

* George W. Bush VS Barack Obama
  * The sample average monthly unemployment rate of Bush is less than that of Obama
  * The two-tailed t-test resulted in a p-value of 2.2e-16
  * A one-tailed t-test where the alternative hypothesis is $\mu_{Bush} < \mu_{Obama}$ results in a p-values of 
  * There exists evidence to reject the null hypothesis which states that there is no statistically significant difference between the average monthly unemployment rates between the Bush and Obama presidential terms.
  * There is also evidence that Bush's monthly average unemployment rates were less than that of Obama's.

## Trend of Unemployment Rates Across Presidential Terms
```{r}
plot_data <- natl_unemploy_pres %>%
  mutate(President = factor(President, levels = unique(pres_data$President)))

ggplot(plot_data, aes(x = DATE, y = UNRATE, color = President)) +
  geom_line() +
  geom_point() +
  labs(title = "Unemployment Rates by President",
       x = "Year",
       y = "Unemployment Rate (%)",
       color = "President") +
  theme_minimal() +
  theme(legend.position = "right") +
  theme(
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14)
  )
```

# Part 3
